Spatio-angular fluorescence microscopy I. Basic theory

### 2018-11-02 papers/basic-theory/paper/paper.tex
Latest draft.

\documentclass{osa-article}

@@

% Talon's custom commands
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rb}{\mathbf{r}_b}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling a microscope imaging an ensemble of in-focus
  fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular transfer function (SATF). We demonstrate our formalism by
  analyzing a paraxial single-view fluorescence microscope without using the
  monopole or scalar approximations. We show that this microscope has an angular
  band limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with numerical phantoms. Notably,
  we show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is imaged in paraxial widefield fluorescence
  microscopes. We discuss the implications of our analysis for all quantitative
  fluorescence microscopy studies and lay out a path towards a complete theory.
\end{abstract*}

\section{Introduction \tb{needs work}}

Fluorescence microscopes are widely used in the biological sciences for
measuring the distribution of fluorophores throughout a sample [cite]. While an
unprocessed fluorescence micrograph reports the approximate distribution of
fluorophores throughout a sample, all microscopes are diffraction limited
[cite], so the image is a blurred version of the true fluorophore distribution.
Therefore, microscopists who are interested in optimal measurements will perform
a computational restoration to recover some of the information lost during the
imaging process.

@@

This work lies at the intersection of three classes of fluorescence microscopy:
(1) single-molecule localization microscopy (SMLM), (2) spatial ensemble
fluorescence microscopy including widefield, confocal, and light-sheet
techniques, and (3) polarized fluorescence microscopy. We briefly review these
three classes and focus on their restoration techniques and use of the monopole
and scalar approximations.

The SMLM community has pioneered the use of rigorous electromagnetic models of
fluorescence microscopes \cite{backer2014, lieb2004} [cite Novotny]. When a
single molecule is fluorescing in the sample, the measured intensity pattern is
strongly dependent on the orientation of the emitting molecule. Backlund and Lew
\cite{backlund2014} have shown that ignoring the orientation of fluorophores can
bias position estimates, so the most accurate SMLM experiments must jointly
estimate the position and orientation of each molecule.

@@

A smaller community of microscopists is interested in measuring the orientation
of ensembles of fluorophores \cite{mehta2016} [Forkey, Goldman, Moerner,
Oldenbourg]. These techniques typically use polarized illumination or polarized
detection to make multiple measurements of the same object. Current angular
restoration techniques use a model of the dipole excitation and emission
processes [Fourkas] to recover the orientation of fluorophores using pixel-wise
arithmetic, but these techniques do not perform any spatial restoration so they
do not use all of the information available to the microscopist. To our
knowledge no work has been done to model the complete spatio-angular response of
fluorescence microscopes to ensembles of oriented fluorophores.

\tb{Out of date}
In this work we model ensembles of in-focus dipole absorber/emitters using
electromagnetic optics and the paraxial approximation. In section 2 we model the
excitation and detection processes of dipoles, and we define the most important
quantity in this work---the spatio-angular transfer function (SATF). We
calculate the SATF and show that fluorescence microscopes have an angular band
limit. In section 3 we perform simulations and compare our model to traditional
scalar models. In section 4 we discuss the implications of our work for
fluorescence microscopy and lay out a path towards a complete spatio-angular
theory of fluorescence microscopy.

\subsection{Monopole emission vs. dipole emission \tb{not sure where this section should live}}
Consider a single dipole radiator at the origin with its dipole moment oriented
along a direction $\so$. The dipole emitter creates a time-harmonic electric
field at a position $\mb{r}$ far from the dipole proportional to
\begin{align}
  \mb{E}_{\text{ff}}(\mb{r}, \so) \propto \frac{\text{exp}[ik|\mb{r}|]}{|\mb{r}|}\, \mh{r}\times\so\times\mh{r}.\label{eq:dip}
\end{align}
A typical approach to deriving Equation \ref{eq:dip} is to (1) use Maxwell's
equations to derive the inhomogeneous electromagnetic wave equations then (2)
use potentials to solve the wave equations with a dipole source term, then (3)
drop the terms that decay faster than $1/|\mb{r}|$ [cite Jackson, Novotny,
Teich]. Equation~\ref{eq:dip} shows that dipole radiators emit spherical
wavefronts of polarized light with amplitude proportional to $\cos\theta$ where
$\theta$ is the angle between the field point $\mb{r}$ and the dipole axis
$\so$.

Many models of fluorophores approximate the vector field on the left hand side
of Equation \ref{eq:dip} with a scalar field---the \textit{scalar
  approximation}---and drop the orientation dependence on the right hand
side---the \textit{monopole approximation}. The approximated field is given by
\begin{align}
  U_{\text{ff}}(\mb{r}) \propto \frac{\text{exp}[ik|\mb{r}|]}{|\mb{r}|}. \label{eq:mono}
\end{align}
Models that use Equation \ref{eq:mono} as a starting point do not account for
polarization or dipole orientation effects.

\section{Theory \tb{outline is okay, subsections vary, recently moved changes of basis to the front}}
We begin our analysis with the abstract Hilbert space formalism of Barrett and
Myers \cite{barrett2004}. Our first task is to formulate the imaging process as
a mapping between two Hilbert spaces $\mc{H}: \mbb{U} \rightarrow \mbb{V}$ where
$\mbb{U}$ is a set than contains all possible objects, $\mbb{V}$ is a set that
contains all possible datasets, and $\mc{H}$ is a model of the instrument that
maps between these two spaces. We denote (possibly infinite-dimensional)
Hilbert-space vectors in $\mbb{U}$ with $\mb{f}$, Hilbert-space vectors in
$\mbb{V}$ with $\mb{g}$, and the mapping between the spaces with

\begin{align}
  \mb{g} = \mc{H}\mb{f}.
\end{align}

Once we have identified the spaces $\mbb{U}$ and $\mbb{V}$, we can start
expressing the mapping between the spaces in a specific object-space and
data-space basis. In most cases the easiest mapping to find uses a
delta-function basis---we expand object and data space into delta functions then
express the mapping as an integral transform. After finding this mapping we can
start to investigate the same mapping in different bases.

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to draw new conclusions about all fluorescence microscopes. In
Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will show that the point spread
function and the optical transfer function are mappings between object and data
space expressed in two different bases. In Section \ref{sec:dipole} we will
extend the monopole imaging model to dipoles and examine the mapping in four
different bases. In Section \ref{sec:para} we will examine the mappings for a
more specific case---a paraxial epi-illumination single-view fluorescence
microscope.

\subsection{Monopole imaging in different bases \tb{50\%}}\label{sec:monopole}
We start by considering a microscope that images a field of in-focus monopoles
by recording the irradiance on a two-dimensional detector. We can represent
the object as a function that assigns a real number to each point on a plane,
so we identify object space as $\mbb{U} = \mbb{L}_2(\mbb{R}^2)$. Similarly, we
have a two-dimensional detector that measures a real number at every point on 
a plane, so data space is the same set $\mbb{V} = \mbb{L}_2(\mbb{R}^2)$. Our
first task is complete---we have identified the geometry of object and data
space.

Next, we find and name the representations of our object and data in a
specific basis. In a delta function basis the object can be represented by a
function $f(\ro)$ called the \textit{monopole density}---the number of
monopoles at the two-dimensional position $\ro$ per unit area. Similarly, in a
delta function basis the data can be represented by a function $g(\rd)$ called
the \textit{irradiance}---the power received by a surface at position $\rd$
per unit area. It may seem pedantic to emphasize that these representations
are in a delta function basis, but we highlight this point to make it clear
that the delta function basis is not special---we can choose a different basis
at will.

% We start by considering a field of in-focus monopoles that have been excited by test
% a spatially uniform beam. As the fluorophores relax, they emit scalar waves in
% the form of Equation \ref{eq:mono}. The waves propagate through the microscope
% and we measure the irradiance on a two dimensional detector.

A reasonable starting point is to assume that the relationship between the
object and the data is \textit{linear}---this is usually true in fluorescence
microscopes since fluorophores emit incoherently so a scaled sum of fluorophores
will result in a scaled sum of the irradiance patterns created by the individual
fluorophores. If the mapping is linear we can write the irradiance as the
integral over a field of monopoles weighted by the detector response due to
single monopoles
\begin{align}
g(\rd) = \int_{\mbb{R}^2}d\ro\, h(\rd,\ro)f(\ro), \label{eq:fwdmono}
\end{align}
where $h(\rd{}, \ro{})$ is the kernel of the integral transform.

Most fluorescence microscopes are (at least approximately) shift-invariant,
which means that shifting the object will result in a corresponding shift of the
data. The mapping between an object and the data in a linear shift-invariant
fluorescence microscope can be written as
\begin{align}
  g(\rd) = \int_{\mbb{R}^2}d\ro\, h(\rd - \ro)f(\ro).  \label{eq:lsi}
\end{align}
When the imaging system is shift invariant the kernel $h(\rd - \ro)$ is usually
called the \textit{point spread function}.

The mapping between the object and the data in linear shift-invariant
microscopes takes a particularly simple form in a spatial harmonic basis. If we
apply the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bs{\nu}) = H(\bs{\nu})F(\bs{\nu}),\label{eq:freq}
\end{align}
where the \textit{irradiance spectrum} is given by
\begin{align}
  G(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
\end{align}
the \textit{optical transfer function} is given by
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
\end{align}
and the \textit{monopole spectrum} is given by
\begin{align}
    F(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}).
\end{align}
Notice that Eqs. \ref{eq:lsi} and \ref{eq:freq} are expressions of the same
mapping between object and data space in different bases. Figure
\ref{fig:monopole-block} summarizes the relationship between object and data
space in both bases.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-block/monopole-block.pdf}
  \caption{We can compute the mapping between object space and data space using
    two different bases.}
     \label{fig:monopole-block}      
\end{figure}

\subsection{Dipole imaging in different bases \tb{50\%}}\label{sec:dipole}
Now we consider a microscope imaging a field of in-focus \textit{dipoles} by
recording the irradiance on a two-dimensional detector. A function that assigns
a real number to each point on a plane is not sufficient to specify a field of
dipoles because the dipoles can have different orientations. To represent the
object we need to extend object space to
$\mbb{U} = \mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ where $\mbb{S}^2$ is the
two-dimensional sphere (the usual sphere embedded in $\mbb{R}^3$).

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{spatio-angular density}---the number of dipoles
at position $\ro{}$ per unit area oriented along $\so{}$ per unit solid angle.
If the imaging system is linear and shift invariant then the mapping between
object and data space can be expressed in a delta function basis as
  \begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so), \label{eq:odpsf}
  \end{align}
  where $h(\rd - \ro, \so)$ is the \textit{orientation-dependent point spread
    function}.

  We can make our first change of basis by applying the Fourier-convolution
  theorem to Eq. \ref{eq:odpsf} which yields
\begin{align}
G(\bs{\nu}) = \int_{\mbb{S}^2}d\mh{s}\, H(\bs{\nu}, \mh{s})F(\bs{\nu}, \mh{s}) \label{eq:odotf},
\end{align}
where we define the \textit{irradiance spectrum} as
\begin{align}
  G(\bs{\nu}) &= \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
  \end{align}
  the \textit{orientation-dependent optical transfer function} as
  \begin{align}
  H(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
  \end{align}
  and the \textit{spatial density spectrum} as
  \begin{align}
  F(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}). 
  \end{align}
  This basis is well suited for simulating objects that are angularly sparse and
  spatially dense.

  The spherical harmonics provide another set of convenient basis functions. We
  can change basis from spherical delta functions to spherical harmonics by
  making use of the generalized Plancharel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}{}\, h(\mh{s})f(\mh{s}) = \lmsum H_\ell^m F_\ell^m, \label{eq:plan}
\end{align}
where
\begin{align}
  F_\ell^m \equiv \int_{\mbb{S}^2}d\mh{s}\, f(\mh{s})\bar{Y}_\ell^m(\mh{s}), 
\end{align}
is the spherical Fourier transform of $f(\mh{s})$, $\bar{z}$ denotes the complex
conjugate of $z$, and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions
defined in Appendix \ref{sec:sph}. Eq. \ref{eq:plan} expresses the fact that
scalar products are invariant under a change of basis (see Eq. 3.78 of
\cite{barrett2004}). The left-hand side of Eq. \ref{eq:plan} is the scalar
product of $\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the
right hand side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a
spherical harmonic function basis. Applying Eq. \ref{eq:plan} to Eq.
\ref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum H_\ell^m(\rd - \ro)F_\ell^m(\ro) \label{eq:atf-form}
\end{align}
where we have defined the \textit{angular transfer function} as
\begin{align}
  H_\ell^m(\mb{r}) = \int_{\mbb{S}^2}d\mh{s}\, h(\mb{r}, \mh{s})\bar{Y}_{\ell}^m(\mh{s}),\label{eq:atf-prep} 
\end{align}
and the \textit{angular density spectrum} as
\begin{align}
  F_\ell^m(\mb{r}) = \int_{\mbb{S}^2}d\mh{s}\, f(\mb{r}, \mh{s})\bar{Y}_{\ell}^m(\mh{s}).
\end{align}
This basis is well suited for simulating objects that are spatially sparse and
angularly dense.

>>>>

\documentclass[]{osa-article}

@@

% Talon's custom commands
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rb}{\mathbf{r}_b}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling a microscope imaging an ensemble of in-focus
  fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular transfer function (SATF). We demonstrate our formalism by
  analyzing a single-view paraxial epi-fluorescence microscope without using the
  monopole or scalar approximations. We show that this microscope has an angular
  band limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with simple phantoms. Notably, we
  show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is recorded by paraxial epi-fluorescence microscopes. We
  discuss the implications of our analysis for all quantitative fluorescence
  microscopy studies and lay out a path towards a complete theory.
\end{abstract*}

\section{Introduction}

Fluorescence microscopes are widely used in the biological sciences for
measuring the spatial distribution of fluorophores throughout a sample [cite].
While an unprocessed fluorescence micrograph reports the approximate
distribution of fluorophores throughout a sample, all microscopes are
diffraction limited [cite], so the image is a blurred version of the true
fluorophore distribution.

@@

The widely-used monopole model treats fluorophores as molecules that absorb and
emit light isotropically. Despite their use in models, monopole
absorber/emitters do not exist in nature. All real fluorophores absorb and emit
light anisotropically, and in this work we investigate the more realistic dipole
model of fluorophores. Since the absorption and emission patterns of many widely
used fluorophores including green fluroescent protein (GFP) are known to be well
described by the dipole model [cite Inoue], we start by considering the dipole
model before higher-order multipole models.

The scalar approximation is another widely-used approximation in fluorescence
microscopy. The scalar approximation models the propagation of light using a
scalar-valued field $U$ instead of the more realistic vector-valued electric
field $\mb{E}$. Although the scalar model yields realistic results in many
cases, the scalar field ignores polarization-dependent effects. The absorption
and emission response of dipoles strongly depends on the polarization of the
fields, so we need to use an electromagnetic model of light propagation to
accurately model a microscope imaging dipoles.

This work lies at the intersection of three classes of fluorescence microscopy:
(1) single molecule localization microscopy (SMLM), (2) spatial ensemble
fluorescence microscopy including widefield, confocal, and light-sheet
techniques, and (3) polarized fluorescence microscopy. We briefly review these
three classes and focus on their use of the monopole and scalar approximations.

The SMLM community has pioneered the use of rigorous electromagnetic models of
fluorescence microscopes \cite{backer2014, lieb2004} [cite Novotny]. When a
single molecule is fluorescing in the sample, the measured intensity pattern is
strongly dependent on the orientation of the emitting molecule. Backlund [cite]
and Lew [cite] \cite{backlund2014} have shown that ignoring the orientation of
fluorophores can bias position estimates, so the most accurate SMLM experiments
must jointly estimate the position and orientation of each molecule.

@@

A smaller community of microscopists is interested in measuring the orientation
of ensembles of fluorophores \cite{mehta2016} [Forkey, Goldman, Moerner,
Oldenbourg]. These techniques typically use polarized illumination or polarized
detection to make multiple measurements of the same object. Current angular
restoration techniques use a model of the dipole excitation and emission
processes [Fourkas] to recover the orientation of fluorophores using pixel-wise
arithmetic, but these techniques do not perform any spatial restoration so they
do not use all of the available information. To our knowledge no work has been
done to model the complete spatio-angular response of fluorescence microscopes
to ensembles of oriented fluorophores.

We begin in Section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently simplify our
analysis of spatio-angular microscopes. In Section \ref{sec:results} we use the
mathematical tools we developed in Section \ref{sec:theory} to analyze and
simulate a single-view paraxial epi-fluorescence microscope, and we demonstrate
our model with four simple phantoms. Finally, in Section \ref{sec:discussion} we
discuss the results and their broader implications.

\section{Theory}\label{sec:theory}
We begin our analysis with the abstract Hilbert space formalism of Barrett and
Myers \cite{barrett2004}. Our first task is to formulate the imaging process as
a mapping between two Hilbert spaces $\mc{H}: \mbb{U} \rightarrow \mbb{V}$, where
$\mbb{U}$ is a set than contains all possible objects, $\mbb{V}$ is a set that
contains all possible datasets, and $\mc{H}$ is a model of the instrument that
maps between these two spaces. We denote (possibly infinite-dimensional)
Hilbert-space vectors in $\mbb{U}$ with $\mb{f}$, Hilbert-space vectors in
$\mbb{V}$ with $\mb{g}$, and the mapping between the spaces with
\begin{align}
  \mb{g} = \mc{H}\mb{f}.
\end{align}

Once we have identified the spaces $\mbb{U}$ and $\mbb{V}$, we can start
expressing the mapping between the spaces in a specific object-space and
data-space basis. In most cases the easiest mapping to find uses a
delta-function basis---we expand object and data space into delta functions,
then express the mapping as an integral transform. After finding this mapping we
can start to investigate the same mapping in different bases.

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to draw new conclusions about all fluorescence microscopes. In
Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will show that the point spread
function and the optical transfer function are mappings between object and data
space expressed in two different bases. In Section \ref{sec:dipole} we will
extend the monopole imaging model to dipoles and examine the mapping in four
different bases.

\subsection{Monopole imaging in different bases}\label{sec:monopole}
We start by considering a microscope that images a field of in-focus monopoles
by recording the irradiance on a two-dimensional detector. This section treads
familiar ground, but it serves to establish the concepts and notation that will
be necessary when we extend to the dipole case.

We can represent the object as a function that assigns a real number to each
point on a plane, so we identify object space as
$\mbb{U} = \mbb{L}_2(\mbb{R}^2)$. Similarly, we have a two-dimensional detector
that measures a real number at every point on a plane, so data space is the same
set $\mbb{V} = \mbb{L}_2(\mbb{R}^2)$.

Next, we name the representations of our object and data in a specific basis. In
a delta function basis the object can be represented by a function $f(\ro)$
called the \textit{spatial monopole density}---the number of monopoles per unit
area at the two-dimensional position $\ro$. Similarly, in a delta function basis
the data can be represented by a function $g'(\rd')$ called the
\textit{irradiance}---the power received by a surface per unit area at position
$\rd'$. It may seem pedantic to emphasize that these representations are in a
delta function basis, but we highlight this point to make it clear that the
delta function basis is not special---we can choose a different basis at will.

A reasonable starting point is to assume that the relationship between the
object and the data is \textit{linear}---this is usually true in fluorescence
microscopes since fluorophores emit incoherently, so a scaled sum of
fluorophores will result in a scaled sum of the irradiance patterns created by
the individual fluorophores. If the mapping is linear, we can write the
irradiance as a weighted integral over a field of monopoles
\begin{align}
g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd',\ro)f(\ro), \label{eq:fwdmono}
\end{align}
where $h'(\rd{}', \ro{})$ is the irradiance at position $\rd'$ created by a
point source at $\ro$.

The irradiance pattern measured by many fluorescence microscopes can be modeled
(at least approximately) as a magnifier with shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro)f(\ro), 
\end{align}
where $m$ is a magnification factor. To simplify our analysis we define a
demagnified detector coordinate $\rd = \rd'/m$ and a normalization factor that
corresponds to the total power incident on the detector plane due to a point
source $P_{\text{mono}} = \int_{\mbb{R}^2}d\mb{r}\,h'(m\mb{r})$. We use these scaling
factors to define the \textit{point spread function} as
\begin{align}
  h(\rd - \ro) = \frac{h'(m[\rd - \ro])}{P_{\text{mono}}},
\end{align}
and the \textit{scaled irradiance} as
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_{\text{mono}}}.
\end{align}
With these definitions we can express the mapping between the object and the
data as a familiar convolution
\begin{align}
  g(\rd) = \int_{\mbb{R}^2}d\ro\, h(\rd - \ro)f(\ro),  \label{eq:lsi}
\end{align}
which will allow us to analyze this class of microscopes without carrying
magnification or normalization factors. We will mostly be using scaled
coordinates, so we have adopted a slightly unusual convention of using primes to
denote unscaled coordinates.

We have chosen to normalize the point spread function so that
\begin{align}
  \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}) = 1. \label{eq:norm}
\end{align}
The point spread function corresponds to a measurable irradiance, so it is
always real-valued and positive.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-block/monopole-block.pdf}
  \caption{The mapping between the object and data space of a monopole
    fluorescence microscope can be computed in two different bases---a delta
    function basis and a spatial harmonic basis. The change of basis is given by
    the two-dimensional Fourier transform denoted $\mathcal{F}_{\mbb{R}^2}$.}
     \label{fig:monopole-block}      
\end{figure}

The mapping between the object and the data in linear shift-invariant
microscopes takes a particularly simple form in a spatial harmonic basis. If we
apply the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bs{\nu}) = H(\bs{\nu})F(\bs{\nu}),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
\end{align}
the \textit{optical transfer function} as
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}).
\end{align}
The point spread function is normalized and real-valued, so we know that the
optical transfer function is normalized $H(0) = 1$ and conjugate-symmetric
$H(-\bs{\nu}) = H^*(\bs{\nu})$.

Notice that Eqs. \ref{eq:lsi} and \ref{eq:freq} are expressions of the same
mapping between object and data space in different bases. Figure
\ref{fig:monopole-block} summarizes the relationship between object and data
space in both bases.

\subsubsection{Monopole coherent transfer functions}

\subsection{Dipole imaging in different bases}\label{sec:dipole}
Now we consider a microscope imaging a field of in-focus dipoles by recording
the irradiance on a two-dimensional detector. A function that assigns a real
number to each point on a plane is not sufficient to specify a field of dipoles
because the dipoles can have different orientations. To represent the object we
need to extend object space to $\mbb{U} = \mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$
where $\mbb{S}^2$ is the two-dimensional sphere (the usual sphere embedded in
$\mbb{R}^3$).

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{spatio-angular dipole density}---the number of
dipoles at position $\ro{}$ per unit area oriented along $\so{}$ per unit solid
angle. Similar to the monopole case, we model the mapping between the object and
the irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so). 
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. Once again, many fluorescence
microscopes can be modeled as magnifiers with shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro, \so)f(\ro, \so). 
\end{align}

We define the same demagnified detector coordinate $\rd = \rd'/m$ and a new
normalization factor that corresponds to the total power incident on the
detector due to a spatial point source with an angularly uniform distribution of
dipoles
$P_\text{dip} = \int_{\mbb{S}^2}d\mb{s}\int_{\mbb{R}^2}d\mb{r}\, h'(m\mb{r},
\mb{s})$. We use these scaling factors to define the
\textit{orientation-dependent point spread function} as
\begin{align}
  h(\rd - \ro, \mb{s}) = \frac{h'(m[\rd - \ro], \mb{s})}{P_\text{dip}}, 
\end{align}
and the \textit{scaled irradiance} as 
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_\text{dip}}. 
\end{align}
With these definitions we can express the mapping between the object and the
data as an integral over the sphere with a spatial convolution at each point
\begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so). \label{eq:odpsf}
\end{align}
Similar to the monopole case, we have chosen to normalize the orientation-dependent point spread function so that
\begin{align}
  \int_{\mbb{S}^2}d\mh{s}\int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \mh{s}) = 1. 
\end{align}
The orientation-dependent point spread function is a measurable quantity, so it
is real-valued and positive.

\subsubsection{Orientation-dependent optical transfer function}
We can make our first change of basis by applying the Fourier-convolution
theorem to Eq. \ref{eq:odpsf} which yields
\begin{align}
G(\bs{\nu}) = \int_{\mbb{S}^2}d\mh{s}\, H(\bs{\nu}, \mh{s})F(\bs{\nu}, \mh{s}) \label{eq:odotf},
\end{align}
where we define the \textit{orientation-dependent optical transfer function} as
  \begin{align}
  H(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
  \end{align}
and the \textit{spatial dipole spectrum} as
  \begin{align}
  F(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}). 
  \end{align}
  Since the orientation-dependent point spread function is normalized and real,
  we know that the orientation-dependent optical transfer function is normalized
  $\int_{\mbb{S}^2}d\mh{s}\, H(0, \mh{s}) = 1$ and conjugate-symmetric
  $H(-\bs{\nu}, \mh{s}) = H^*(\bs{\nu}, \mh{s})$.

  This choice of basis is well suited for simulating and analyzing objects that
  are angularly sparse and spatially dense.

  \begin{figure}
  \hspace{-5em}
  \includegraphics[scale=1.0]{../figures/dipole-block/dipole-block.pdf}
  \caption{The mapping between the object and data space of a dipole
    fluorescence microscope can be computed in four different bases---a delta
    function basis, a spatial-harmonic/angular-delta basis, a
    spatial-delta/spherical-harmonic basis, and a spatio-angular harmonic basis.
    The changes of basis are given by the two-dimensional Fourier transform
    denoted $\mathcal{F}_{\mbb{R}^2}$ and the spherical Fourier transform
    denoted $\mathcal{F}_{\mbb{S}^2}$.}
   \label{fig:dipole-block}      
    \end{figure}
